On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-10 manuscripts/PNAS/chazthm_pnas.tex
lil edits

Here, we prove very generally that uniqueness and stability in sparse linear coding is an expected property of the model. %; that is, it is a well-posed problem (in the sense of Hadamard \cite{Hadamard1902}). 
%In other words, whenever enough sparsely represented data generated from a dictionary (satisfying a spark condition) are observed, the original codes and dictionary are uniquely determined up to an error that is commensurate with the measurement noise. 
More specifically, dictionaries are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq k$, in almost all cases the dictionary learning problem is well-posed (in the sense of Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies a spark condition. Our results  also apply when only an upper bound on the number $m$ of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of dictionary learning.

@@

Since $f$ is an analytic function, having one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set of measure zero. Fortunately, such a matrix $\mathbf{A}$ is easily constructed by adding rows of zeroes to any $k \times m$ ``Vandermonde'' matrix $[\gamma_i^j]_{i,j=1}^{k,m}$ with distinct $\gamma_i$ (so that each term in the product above is nonzero).
%$A_{ij} = \gamma_i (since the $\gamma_i$ are distinct, any $k \times k$ ``Vandermonde'' sub-determinant is nonzero)
%inserting $n-k$ rows of zeros into a $k \times m$ ``Vandermonde" matrix construction. %Let $\gamma_1, \ldots, \gamma_m$ be any distinct numbers. Then the columns of the $k \times m$ matrix $V = (\gamma^{\ell}_i)^{k,m}_{\ell,i=1}$ are in general linear position (since the $\gamma_i$ are distinct, any $k \times k$ ``Vandermonde" sub-determinant is nonzero). Concatenating $n-k$ rows of zeros to the bottom of this matrix yields a $n \times m$ matrix $\mathbf{A}$ satisfying the spark condition, i.e. with $f(\mathbf{A}) = 0$.
Hence, almost every real matrix $\mathbf{A}$ with at least $k$ rows and columns satisfies \eqref{SparkCondition}.

@@

In fact, though, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have utilized compressed sensing for signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of CS, it is only a matter of time before these systems incorporate sparse representational learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) are equivalent as soon as enough data originate from a statistically identical system.
%Data science:
%  
%in a common application of compressed sensing in engineering

@@

%\begin{lemma}\label{NonEmptyLemma} 
%Fix $\ell, m$, and $\bar m$. Suppose $E \subseteq 2^{[m]}$ is $\ell$-regular with the SIP.  If $\pi: E \to 2^{[\bar m]}$ has $\sum_{S \in E} |S| = \sum_{S \in E} |\pi(S)|$ and:
%\begin{align}\label{cond}
%|\bigcap_{S \in F} \pi(S)| \leq |\bigcap_{S \in F} S |,\ \ \   \text{for } F \subseteq E,
%\end{align}
%then $\bar m \geq m$ and the association $i \mapsto \cap \pi(F(i))$ defines an injective map from $J$ to $[\bar m]$ for some $J \subseteq [m]$ of size $p = \bar m - \ell(\bar m - m)$. In particular, if $\bar m = m$ then the map $\pi$ is induced by a permutation.
%\end{lemma}
% i'm placing this on hold since one thing i found was that i needed pi to be injective, but for that i needed the fact that |pi(S)| = |S|

\begin{lemma}\label{NonEmptyLemma} 
Fix $\ell, m$, and $\bar m$. Suppose $E \subseteq 2^{[m]}$ is $\ell$-regular with the SIP.  If $\pi: E \to 2^{[\bar m]}$ has $|S| = |\pi(S)|$ and:
\begin{align}\label{cond}
|\bigcap_{S \in F} \pi(S)| \leq |\bigcap_{S \in F} S |,\ \ \   \text{for } F \subseteq E,
\end{align}
then $\bar m \geq m$ and the association $i \mapsto \cap \pi(F(i))$ defines an injective map from $J$ to $[\bar m]$ for some $J \subseteq [m]$ of size $p = \bar m - \ell(\bar m - m)$. In particular, if $\bar m = m$ then the map $\pi$ is induced by a permutation.
\end{lemma}

\begin{proof}
First observe that assumption \eqref{cond} already constrains $\pi$ to be injective:  if $S_1 \neq S_2$, then $|\pi(S_1) \cap \pi(S_2)| \leq |S_1 \cap S_2| < \max\{|S_1|, |S_2|\} = \max\{|\pi(S_1)|, |\pi(S_2)|\}$, which implies that $\pi(S_1) \neq \pi(S_2)$.  Next, consider the collection of pairs: \[T_1 := \{(i, S): i \in \pi(S), S \in E\},\] which number $|T_1| = \sum_{S \in E} |\pi(S)| = \sum_{S \in E} |S| = \ell m$ by regularity of $E$.  
Our second observation is that $\bar m \geq m$. Otherwise, if $\bar m < m$, then by pigeonholing the $m \ell \geq \ell (\bar m + 1) \geq [(\ell + 1) - 1] \bar m + 1$ elements of $T_1$ with respect to the set $[\bar m]$ of possible first indices, there must exist at least $\ell + 1$ elements of $T_1$ having the same first index. But by \eqref{cond} and injectivity of $\pi$, we then have more than $\ell$ edges in $E$ with common element, a contradiction to regularity.

Since $|T_1| > \bar m (\ell -1) + (p-1) \geq \bar m (\ell -1)$, the pigeonhole principle also implies that there are at least $\ell$ elements of $T_1$ sharing the same first coordinate. Let $F_1$ be the preimage of this family under $\pi$.  By the assumption \eqref{cond}, the intersection of $F_1$ is nonempty, and in fact must contain exactly $\ell$ elements. Thus, $F_1$ is the star of some element $i_1 \in [m]$.  From the SIP, the intersection of sets in $F_1$ is $\{i_1\}$, which by \eqref{cond} implies that the intersection of $\pi(S)$ with $S \in F_1$ also contains exactly one element.  If $p > 1$, then define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains more than $(\bar m - 1)(\ell -1)$ sets, and produce by the same argument a (necessarily) different index $i_2$. Iterating $p = \bar{m} - \ell(\bar m - m)$ times this loop yields the desired $J = \{i_1, \ldots, i_p\} \subseteq [m]$.
\end{proof}
\begin{proof}[Proof of Lem.\ref{MainLemma}]
We begin by showing that: 
\begin{align}\label{eq2}
d(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}, \text{\rm Span}\{\mathbf{A}_S\} ) = d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}).
\end{align}
Since the right-hand side of \eqref{GapUpperBound} is strictly less than one (by \eqref{delrho} and $r \leq 1$), it follows by \eqref{dimLem} that $|\pi(S)| \geq \dim(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) \geq \dim(\text{\rm Span}\{\mathbf{A}_{S}\}) = |S|$, with the equality due to $\mathbf{A}$ satisfying \eqref{SparkCondition}. Since $|S| = |\pi(S)|$, we therefore have $\dim(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) = \dim(\text{\rm Span}\{\mathbf{A}_{S}\})$, and \eqref{eq2} follows by \eqref{eqdim}. Note that the columns of $\mathbf{B}_{\pi(S)}$ are therefore linearly independent, for all $S \in E$.

We next verify \eqref{cond} holds.  Fix $F \subseteq E$. Since $\text{\rm Span}\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\} \subseteq \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}$, if $\cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\} = \{\textbf{0}\}$ then we must have $|\cap_{S \in F} \pi(S)| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} trivially holds. Suppose then that the intersection is not the zero vector. By Lem.~\ref{SpanIntersectionLemma} and Lem.~\ref{DistanceToIntersectionLemma}, and incorporating \eqref{eq2} and \eqref{GapUpperBound}, we have:
\begin{align}\label{randoml}
d( \text{\rm Span}&\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\}, \text{\rm Span}\{\mathbf{A}_{\cap_{S \in F} S}\}  ) \nonumber \\
&\leq d\left( \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}, \cap_{S \in F} \text{\rm Span}\{\mathbf{A}_{S}\} \right) \nonumber \\
&\leq \sum_{T \in F} \frac{ d\left( \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\},\text{\rm Span}\{\mathbf{A}_{T}\} \right) }{ r( \{ \text{\rm Span}\{\mathbf{A}_{S}\}_{S \in F}) } \nonumber \\
&\leq \sum_{T \in F} \frac{ d\left( \text{\rm Span}\{\mathbf{B}_{\pi(T)}\},\text{\rm Span}\{\mathbf{A}_{T}\} \right) }{ r( \{ \text{\rm Span}\{\mathbf{A}_{S}\}_{S \in F}) }\nonumber \\
&\leq \frac{|F| \varepsilon}{r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})} 
\leq \frac{C_2 \varepsilon}{\max_i\|\mathbf{A}_i\|_2}. 
\end{align}
Since $\varepsilon < L_2(\mathbf{A}) / C_2$, by \eqref{delrho} the right-hand side in \eqref{randoml} is strictly less than one. Hence, $\dim(\text{\rm Span}\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\}) \leq \dim(\text{\rm Span}\{\mathbf{A}_{\cap_{S \in F} S}\})$ by \eqref{dimLem}, and \eqref{cond} follows from the linear independence of the columns of $\mathbf{A}_{S}$ and $\mathbf{B}_{\pi(S)}$ for all $S \in F$.

>>>>

Here, we prove very generally that uniqueness and stability in sparse linear coding is an expected property of the model. %; that is, it is a well-posed problem (in the sense of Hadamard \cite{Hadamard1902}). 
%In other words, whenever enough sparsely represented data generated from a dictionary (satisfying a spark condition) are observed, the original codes and dictionary are uniquely determined up to an error that is commensurate with the measurement noise. 
More specifically, dictionaries are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq \max(2k,m)$, in almost all cases the dictionary learning problem is well-posed (in the sense of Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies a spark condition. Our results also apply when the number $m$ of dictionary elements is unknown. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of dictionary learning.

@@

Since $f$ is an analytic function, having one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set of measure zero. Fortunately, such a matrix $\mathbf{A}$ is easily constructed by adding rows of zeroes to any $\max(2k,m) \times m$ ``Vandermonde'' matrix $[\gamma_i^j]_{i,j=1}^{k,m}$ with distinct $\gamma_i$ (so that each term in the product above is nonzero).
%$A_{ij} = \gamma_i (since the $\gamma_i$ are distinct, any $k \times k$ ``Vandermonde'' sub-determinant is nonzero)
%inserting $n-k$ rows of zeros into a $k \times m$ ``Vandermonde" matrix construction. %Let $\gamma_1, \ldots, \gamma_m$ be any distinct numbers. Then the columns of the $k \times m$ matrix $V = (\gamma^{\ell}_i)^{k,m}_{\ell,i=1}$ are in general linear position (since the $\gamma_i$ are distinct, any $k \times k$ ``Vandermonde" sub-determinant is nonzero). Concatenating $n-k$ rows of zeros to the bottom of this matrix yields a $n \times m$ matrix $\mathbf{A}$ satisfying the spark condition, i.e. with $f(\mathbf{A}) = 0$.
Hence, almost every real $n \times m$ matrix $\mathbf{A}$ with $n \geq \max(2k,m)$ satisfies \eqref{SparkCondition}.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have utilized compressive sensing for signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressive sensing, it is only a matter of time before these systems incorporate sparse representational learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) are equivalent as soon as enough data originate from a statistically identical system.
%Data science:
%  
%in a common application of compressed sensing in engineering

@@

\begin{lemma}\label{NonEmptyLemma} 
Fix $\ell, m$, and $\bar m$. Suppose $E \subseteq 2^{[m]}$ is $\ell$-regular with the SIP.  If $\pi: E \to 2^{[\bar m]}$ has $\sum_{S \in E} |S| = \sum_{S \in E} |\pi(S)|$ and:
\begin{align}\label{cond}
|\bigcap_{S \in F} \pi(S)| \leq |\bigcap_{S \in F} S |,\ \ \   \text{for } F \subseteq E,
\end{align}
then $\bar m \geq m$ and the association $i \mapsto \cap \pi(F(i))$ defines an injective map from $J$ to $[\bar m]$ for some $J \subseteq [m]$ of size $p = \bar m - \ell(\bar m - m)$. In particular, if $\bar m = m$ then the map $\pi$ is induced by a permutation.
\end{lemma}

\begin{proof}
Consider the collection of pairs: \[T_1 = \{(i, S): i \in \pi(S), S \in E\},\] which number $|T_1| = \sum_{S \in E} |\pi(S)| = \sum_{S \in E} |S| = \ell m$ by regularity of $E$.  
We first show that $\bar m \geq m$. Otherwise, if $\bar m < m$, then by pigeonholing the $m \ell \geq \ell (\bar m + 1) \geq [(\ell + 1) - 1] \bar m + 1$ elements of $T_1$ with respect to the set $[\bar m]$ of possible first indices, we see that there must exist at least $\ell + 1$ elements of $T_1$ having the same first index. But by \eqref{cond} there can be no more than $\ell$ elements of $T_1$ with a given first index, since $E$ is $\ell$-regular. Hence $\bar m \geq m$.

Next, since \[|T_1| > \bar m (\ell -1) + (p-1) > \bar m (\ell -1),\] the pigeonhole principle also implies that there are at least $\ell$ elements of $T_1$ sharing the same first coordinate. Let $F_1$ be the preimage of this family under $\pi$.  By the assumption \eqref{cond}, the intersection of $F_1$ is nonempty, and in fact must contain exactly $\ell$ elements. Thus, $F_1$ is the star of some element $i_1 \in [m]$.  From the SIP, the intersection of sets in $F_1$ is $\{i_1\}$, which by \eqref{cond} implies that the intersection of $\pi(S)$ with $S \in F_1$ also contains exactly one element.  Set $T_2 = T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains at least $(\bar m - 1)(\ell -1)$ sets. Iterating this argument $p = \bar{m} - \ell(\bar m - m)$ times produces the desired $J = \{i_1, \ldots, i_p\} \subseteq [m]$.
\end{proof}
\begin{proof}[Proof of Lem.\ref{MainLemma}]
We begin by showing that: 
\begin{align}\label{eq2}
d(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}, \text{\rm Span}\{\mathbf{A}_S\} ) = d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}).
\end{align}
Since the right-hand side of \eqref{GapUpperBound} is strictly less than one (by \eqref{delrho} and $r \leq 1$), it follows by \eqref{dimLem} that $|\pi(S)| \geq \dim(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) \geq \dim(\text{\rm Span}\{\mathbf{A}_{S}\}) = |S|$, with the equality due to $\mathbf{A}$ satisfying \eqref{SparkCondition}. Since $|S| = |\pi(S)|$, we therefore have $\dim(\text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) = \dim(\text{\rm Span}\{\mathbf{A}_{S}\})$, and \eqref{eq2} follows by \eqref{eqdim}. Note that the columns of $\mathbf{B}_{\pi(S)}$ are therefore linearly independent, for all $S \in E$.

We next verify \eqref{cond} holds.  Fix $F \subseteq E$. Since $\text{\rm Span}\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\} \subseteq \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}$, if $\cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\} = \textbf{0}$ then we must have $|\cap_{S \in F} \pi(S)| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} trivially holds. Suppose then that the intersection is not the zero vector. By Lem.~\ref{SpanIntersectionLemma} and Lem.~\ref{DistanceToIntersectionLemma}, and incorporating \eqref{eq2} and \eqref{GapUpperBound}, we have:
\begin{align}\label{randoml}
d( \text{\rm Span}&\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\}, \text{\rm Span}\{\mathbf{A}_{\cap_{S \in F} S}\}  ) \nonumber \\
&\leq d\left( \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}, \cap_{S \in F} \text{\rm Span}\{\mathbf{A}_{S}\} \right) \nonumber \\
&\leq \sum_{T \in F} \frac{ d\left( \cap_{S \in F} \text{\rm Span}\{\mathbf{B}_{\pi(S)}\},\text{\rm Span}\{\mathbf{A}_{T}\} \right) }{ r( \{ \text{\rm Span}\{\mathbf{A}_{S}\}_{S \in F}) } \nonumber \\
&\leq \sum_{T \in F} \frac{ d\left( \text{\rm Span}\{\mathbf{B}_{\pi(T)}\},\text{\rm Span}\{\mathbf{A}_{T}\} \right) }{ r( \{ \text{\rm Span}\{\mathbf{A}_{S}\}_{S \in F}) }\nonumber \\
&\leq \frac{|F| \varepsilon}{r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})} 
\leq \frac{C_2 \varepsilon}{\max_i\|\mathbf{A}_i\|_2}. 
\end{align}
Since $\varepsilon < L_2(\mathbf{A}) / C_2$, by \eqref{delrho} the right-hand side in \eqref{randoml} is strictly less than one. Hence, $\dim(\text{\rm Span}\{\mathbf{B}_{\cap_{S \in F}\pi(S)}\}) \leq \dim(\text{\rm Span}\{\mathbf{A}_{\cap_{S \in F} S}\})$ by \eqref{dimLem}, and \eqref{cond} follows from the linear independence of the columns of $\mathbf{A}_{S}$ and $\mathbf{B}_{\pi(S)}$ for all $S \in F$.
